---
title: "SEIRD contact matrix"
output: rmarkdown::pdf_document
bibliography: ../references.bib
vignette: >
  %\VignetteIndexEntry{SEIRD_contact_matrix}
  %\VignetteEncoding{UTF-8}
  %\VignetteEngine{knitr::rmarkdown}
editor_options:
  markdown:
    wrap: 72
---

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>"
)
```

```{r setup}
library(comomodels)
library(tidyverse)
library(ggplot2)
library(socialmixr)
```

## Introduction

Age-structured compartmental models such as the SEIRD implemented in comomodels
use contact matrices to specify the spread of a disease within and between age
groups. Given a contact matrix $C$, each element $C_{i,j}$ indicates the
expected number of contacts someone from age group $i$ has per day with people
from age group $j$.

Comomodels includes estimates of the contact matrix for each country ([@prem2021projecting]; 
with full details available in the data documentation). Separate
matrices are available for contacts at home, work, school, and other. Below, we
generate a plot of the contact matrices.

```{r}
contact_home <- comomodels::contact_home
contact_work <- comomodels::contact_work
contact_school <- comomodels::contact_school
contact_other <- comomodels::contact_other
population <- comomodels::population
```

```{r, fig.width=10, fig.height=7}
# reformat matrices for plotting
ages <- seq(0, 80, 5)
age_names <- vector(length = 16)
for(i in seq_along(age_names)) {
  age_names[i] <- paste0(ages[i], "-", ages[i + 1])
}

format_matrix <- function(contact_matrix, age_names) {
  colnames(contact_matrix) <- age_names
  contact_matrix$age_infectee <- age_names
  contact_matrix %>%
    pivot_longer(all_of(age_names)) %>% 
    rename(age_infector=name) %>% 
    mutate(age_infector=fct_relevel(age_infector, age_names)) %>% 
    mutate(age_infectee=fct_relevel(age_infectee, age_names))
}

c_home <- format_matrix(contact_home$"United Kingdom", age_names) %>% mutate(type="home")
c_work <- format_matrix(contact_work$"United Kingdom", age_names) %>% mutate(type="work")
c_school <- format_matrix(contact_school$"United Kingdom", age_names) %>% mutate(type="school")
c_other <- format_matrix(contact_other$"United Kingdom", age_names) %>% mutate(type="other")

c_all <- c_home %>%
  bind_rows(c_work) %>% 
  bind_rows(c_school) %>% 
  bind_rows(c_other)

# plot all
c_all %>%
  ggplot(aes(x=age_infector, y=age_infectee, fill=value)) +
  geom_tile() +
  scale_fill_viridis_c() +
  facet_wrap(~type) +
  theme(axis.text.x = element_text(angle=90, vjust=0.5, hjust=1))
```

While it is possible to construct location-based transmission models, in this 
study we consider only age structure. Thus, we obtain the total contact matrix
for the age-structured SEIRD model by summing the four location specific
contact matrices. Further details, including an explanation of the model 
equations, can be found in the SEIRD_age_structured vignette in the comomodels
package:
https://github.com/Como-DTC-Collaboration/como-models/blob/main/vignettes/SEIRD_age_structured.Rmd

## Uncertainty in the contact matrix

Typically, when performing simulations of the age-structured SEIR model, or
inference for the parameters of the model, the contact matrix is provided as a
fixed input. However, using fixed values for the contact matrix neglects the
uncertainty which may be present in the contact data.

The purpose of the remainder of this notebook is to investigate the sensitivity
of the outputs of the age-structured SEIRD model to the values of the contact
matrix. Using bootstrap samples which represent the uncertainty in the contact
matrix, we show significant uncertainty in the numbers of infected individuals.
The bootstrap algorithm works by repeatedly selecting random samples (with
replacement) of the survey respondents, and using the sample to calculate a
particular contact matrix. Each sample yields a potentially different contact
matrix, and this set of contact matrices reveals something of the uncertainty
in its values.

## Accuracy of uncertainty estimates produced by the bootstrapping methods

The bootstrap algorithm is an easy way to obtain some idea of the uncertainty in
the contact matrix. However, due to the simplicity of the procedure, its
results should be treated as mere approximations of the true uncertainty that
may exist in the contact matrix. In particular, the bootstrap algorithm does 
not account for the possibility that the contact data in the original survey is
unrepresentative of the actual population. For example, if contact data is 
collected primarily from an urban area in a country whose population is mostly
rural, the resulting contact matrices may be inaccurate for the country, and 
the bootstrap cannot account for the lack of information in the original data.

## Bootstrap samples of contact matrices

We obtain samples of the contact matrix using the socialmixr library which
accesses the POLYMOD data. This library allows us to generate bootstrap samples
of the contact matrix for countries covered by the study. The first step is to
generate 200 of these samples for the contact matrix in the United Kingdom.

```{r}

# Define age groups and names
ages <- seq(0, 80, 5)
age_names <- vector(length = 16)
for(i in seq_along(age_names)) {
  age_names[i] <- paste0(ages[i], "-", ages[i + 1])
}

# Get population data
pops <- population[population$country == "United Kingdom", ]$pop
pop_fraction <- pops/sum(pops)
pop_fraction[16] <- sum(pop_fraction[16:21])
pop_fraction <- pop_fraction[1:16]
n_ages <- 16

# Load the contact matrix data from POLYMOD and get bootstrap samples
n_bootstrap <- 200
data(polymod)
polymod_data <- contact_matrix(polymod,
                               n=n_bootstrap,
                               countries="United Kingdom",
                               age.limits=ages)

# Get the first element of the list, which contains the matrices
matrices <- polymod_data["matrices"][[1]]

```

First, we inspect the range of values in the sampled contact matrices. In the
plot below, we look at the distribution of the diagonal elements of the matrix
for each age group.

```{r, fig.width=7, fig.height=5}

contacts_same_age <- c()
ages_list <- c()
for (i in 1:n_bootstrap){
  contacts_same_age <- append(contacts_same_age, diag(matrices[[i]][[1]])[1:16])
  ages_list <- append(ages_list, age_names)
}

data <- data.frame(ages_list, contacts_same_age)
data$ages_list <- factor(data$ages_list, levels=age_names[1:16], ordered=TRUE)

ggplot(data, aes(x=ages_list, y=contacts_same_age), l) + geom_boxplot() +
  xlab("Age group") + ylab("Number of contacts with same age group")
```

The plot shows the most uncertainty for ages 5--20 (schoolchildren). In most
of the other age groups, the uncertainty is small but nowhere does it appear 
negligible. In the next section, the effect of these uncertainties on the
SEIR outputs will be studied.

## SEIRD simulations

In this step, we run the age-structured SEIRD model once for each bootstrap
sample of the contact matrix. We use fixed values for the other parameters of
the model. We set an initial exposed group compartment equal to 0.1% of the
population size, and all other individuals in the susceptible group.

The transmission parameters are set to resemble those of the COVID-19 outbreak.
For the latency period, we assume 5.5 days [@xin2021estimating], and for the duration
of infectiousness, 2 days (thus, fewer than 1% of individuals are infectious after
10 days, in line with government guidelines [@stayathome2021]). To set the 
transmission rate $\beta$, we choose the value such that the basic reproduction 
number, $R_0$, is equal to 2.4 [@ferguson2020report]. The age-structured model also takes the parameter $\mu$ which specifies the rate
at which infected people enter the deceased compartment. For this parameter, we
provide a vector giving a separate value for each age group. The values are
selected to resemble the death rates of the COVID-19 outbreak in mainland China
during January and February 2020 [@verity2020estimates]. To obtain estimates of
$\mu$ for each age group, we use data for the case fatality ratio (CFR).

To calculate the recovery rate ($\gamma$) and the death rate ($\mu$) from the above quantities, we introduce $\eta$ giving the
inverse of the average duration of infectiousness, and set:

$$\mu = \eta \text{CFR}$$
$$\gamma = \eta (1-\text{CFR})$$



```{r}
cfr_vals = c(0.00260, 0.00260, 0.0148, 0.0148, 0.0600, 0.0600, 0.146, 0.146, 
             0.295, 0.295, 1.25, 1.25, 3.99, 3.99, 8.61, 8.61)
eta = 0.5
gamma_age_vals <- eta * (1 - cfr_vals / 100)
mu_age_vals <- eta * cfr_vals / 100
```

```{r}
for (i in 1:n_bootstrap){
  matrix=matrices[[i]][[1]]

  # Remove the column and row names so the model will accept it
  colnames(matrix) <- NULL
  rownames(matrix) <- NULL

  # Keep the data for ages 0-80, in 5 year increments
  matrix <- matrix[1:16,1:16]

  model <- comomodels::SEIRDAge(n_age_categories=n_ages,
                   contact_matrix=matrix,
                   age_ranges=as.list(age_names))

  # Set the other parameters of the model
  transmission_parameters(model) <- list(b=1.64, k=1/5.5, g=gamma_age_vals, mu=mu_age_vals)
  initial_conditions(model) <- list(S0=pop_fraction*0.999,
                                    E0=rep(0, n_ages),
                                    I0=pop_fraction*0.001,
                                    R0=rep(0, n_ages),
                                    D0=rep(0, n_ages))
  res <- run(model, time=seq(0, 100, by=1))
  
  # Get states from results
  res <- res[['states']]

  # Save the data for the I and R compartments
  x = filter(res, compartment %in% c("I", "R", "D"))
  if (i==1)
    all_results <- x
  else
    all_results <- rbind(all_results, x)
}
```

Having obtained the simulation results, we plot the central 90% probability
interval of the number in the infected compartment over time for two selected
age groups (15--20 and 75--80).


```{r, fig.width=7, fig.height=5}

I_data <- filter(all_results, age_range=="15-20", compartment=="I")
data <- I_data$value * sum(pops)
dim(data) <- c(length(I_data$time)/n_bootstrap, n_bootstrap)
quants <- t(apply(data, 1, quantile, probs=c(0.05, 0.5, 0.95), na.rm=TRUE))
quants_df <- data.frame(quants)
quants_df["time"] <- seq(0, 100, by=1)

ggplot(quants_df, aes(x = time)) +
  geom_line(aes(y=X50.)) +
  geom_ribbon(aes(ymin=X5., ymax=X95.), fill="blue", alpha=0.5) +
  ylab("I compartment, age 15-20")

I_data <- filter(all_results, age_range=="75-80", compartment=="I")
data <- I_data$value * sum(pops)
dim(data) <- c(length(I_data$time)/n_bootstrap, n_bootstrap)
quants <- t(apply(data, 1, quantile, probs=c(0.05, 0.5, 0.95), na.rm=TRUE))
quants_df <- data.frame(quants)
quants_df["time"] <- seq(0, 100, by=1)

ggplot(quants_df, aes(x = time)) +
  geom_line(aes(y=X50.)) +
  geom_ribbon(aes(ymin=X5., ymax=X95.), fill="blue", alpha=0.5) +
  ylab("I compartment, age 75-80")

```

The results show that the while the shape of the epidemic trajectory remains
similar for all contact matrices in the bootstrap sample, the number of people
in the infected compartment exhibits significant uncertainty, particularly near the
peak of the epidemic.

Next, we perform the uncertainty analysis for the number in the recovered
compartment at the final time point, for all age groups.


```{r, fig.width=7, fig.height=5}

data <- filter(all_results, compartment=="R", time==100.0)

ggplot(data, aes(x=age_range, y=value/pop_fraction), l) +
  geom_boxplot() +
  xlab("Age group") +
  ylab("Proportion of age group ever infected")

```
These results show that younger age groups suffer more infections. This trend 
can be explained by the larger numbers of total contacts for younger age
groups, as observed in the contact matrices at the beginning of this notebook.

Finally, we study the effects on death.

```{r, fig.width=7, fig.height=5}

data <- filter(all_results, compartment=="D", time==100.0)

ggplot(data, aes(x=age_range, y=value/pop_fraction), l) +
  geom_boxplot() +
  xlab("Age group") +
  ylab("Proportion of age group died")

```
Although younger people are more likely to be infected in this simulation,
deaths occur mainly in the elderly, due to the age-structured mortality
parameter $\mu$ described above. The different bootstrap samples of the contact
matrix result in a wide range in the number of deaths, particularly in the
70--80 age groups.
