---
title: "On the numerical solution of compartmental models"
output: rmarkdown::pdf_document
bibliography: ../references.bib
vignette: >
  %\VignetteIndexEntry{Numerical solution of SEIRD}
  %\VignetteEncoding{UTF-8}
  %\VignetteEngine{knitr::rmarkdown}
editor_options:
  markdown:
    wrap: 72
---

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>",
  warning = FALSE,
  message = FALSE
)
```



## Introduction
In comomodels, we focus on deterministic models which can be used to represent infection dynamics. Specifically, we investigate compartmental models, which are described by systems of _ordinary differential equations_. In these models, the rate of evolution of the state variables through time is modelled as a function of time, parameters and the current state values. If these differential equations are coupled with a set of initial conditions, they form a system whose solution represents the evolution of the state variables over time. Like many differential equation models of interest in scientific
applications, typical compartmental models are nonlinear, and they generally cannot be solved exactly. Instead, an approximation of the solution must be obtained using numerical methods, which, in effect, numerically integrate the system by taking small discrete time steps.

Systems of differential equations can be integrated using a fixed time step solver, such as the _forward Euler_ method. In this notebook, we show that when such methods are used without sufficiently small time steps, the numerical solution can represent a poor approximation. We recommend using more sophisticated adaptive step size solvers to integrate the types of (deterministic) compartmental models encountered in infectious disease epidemiology, as are implemented as defaults in the comomodels package.

This notebook begins with a review of the use of forward Euler on a toy problem with a known solution. Then, it demonstrates the importance of accurate ODE solvers for comomodel's `SEIRD` model.


```{r setup}
library(comomodels)
library(deSolve)
library(ggplot2)
library(tidyverse)
library(glue)
```

## Numerical solution of a differential equation using forward Euler

One of the simplest numerical solvers for differential equations is the forward Euler method. This method assumes a first order differential equation:

$$\frac{df}{dt} = g(t, f)$$

(note that $f$ may be a vector-valued function), as well as an initial condition $f(t=t_0)$. Forward Euler forms an approximate solution $\{f_i\}_{i=0}^N$ on a set of time points 
$\{t_i = t_0 + i \Delta t\}_{i=0}^N$ which is given by:

$$f_0 = f(t=t_0);$$
$$f_{i+1} = f_i + g(t_i, f_i) \Delta t,\:\:\:\: i=1, \dots, N-1.$$

### Importance of the solver step size

Note that $\Delta t$—the time spacing between adjacent values in the 
approximate solution—is a critical parameter of the forward Euler method. While 
calculating the next function value $f_{i+1}$, this method approximates 
the derivative $g(t, f)$ with the constant value $g(t_i, f_i)$ for the interval
$(t_i, t_i + \Delta t)$. However, $g(t, f)$ typically varies
over time—thus, as $\Delta t$ increases, this approximation becomes worse, and this approximation error accumulates in the numerical solution and causes it to deviate from the true values.

However, obtaining a satisfactory solution to the differential equation is not
as simple as setting $\Delta t$ to a tiny value, as decreasing the 
value of $\Delta t$ increases the runtime of the solver, and for
very small values of $\Delta t$, the runtime may be prohibitive. This is because solving the ODE on
a time interval of length $T$ requires $T/\Delta t$ iterations of the forward Euler method: each
iteration requiring one evaluation of the function $g$—smaller values of $\Delta t$ thus 
require more calculations of $g$, which are potentially slow.

### Effect of solver step size on solution accuracy

We examine the numerical error arising from the solver step size in the SIR model. The 
proportions of susceptible (S), infected (I), and recovered (R) are modelled according to the differential equations:

$$\frac{dS}{dt} = -\beta S I,$$

$$\frac{dI}{dt} = \beta S I - \gamma I,$$
and 
$$\frac{dR}{dt} = \gamma I. $$

We focus on the $I$ compartment and consider the early phase of an epidemic where $S>> I$.
Under these conditions, $S$ is approximately constant and the equation for infectious individuals is given by:

$$\frac{dI}{dt} = (\beta - \gamma) S I;\;\;\;\; I(t=0) = I_0$$

whose analytical solution is:

$$I(t) = I_0 \exp((\beta - \gamma) S t) $$

We set $I_0 = 0.001$ and $(\beta - \gamma) S = 1$. We solve this differential equation on
the interval from $t=0$ to $t=3$ with two different choices of the time step -- $\Delta t=1$ day and $\Delta t=0.1$ day -- and in both cases, we compare the obtained approximate solutions with the known true solution.

```{r, fig.width=7, fig.height=5}
rhs <- function(t, state, params){
  list(state)
}

true_solution <- function(t){
  0.001 * exp(t)
}


sparse_times <- seq(0, 3, by=1)
dense_times <- seq(0, 3, by=0.1)


# Solve using "euler" from deSolve
# This implementation of the method uses the same times on which output is
# requested as the solver time points
numerical_solution_sparse = ode(
  y=0.001,
  times=sparse_times,
  parms=NULL,
  func=rhs,
  method="euler",
)

numerical_solution_dense = ode(
  y=0.001,
  times=dense_times,
  parms=NULL,
  func=rhs,
  method="euler",
)


# Save the results in data frames for plotting
results_sparse <- data.frame(
  t=numerical_solution_sparse[,1],
  f=numerical_solution_sparse[,2],
  method="Numerical, dt=1"
)

results_dense <- data.frame(
  t=numerical_solution_dense[,1],
  f=numerical_solution_dense[,2],
  method="Numerical, dt=0.1"
)

results_true <- data.frame(
  t=seq(0, 3, by=0.01),
  f=true_solution(seq(0, 3, by=0.01)),
  method="True solution"
)


ggplot() +
  geom_point(data=rbind(results_sparse, results_dense),
             aes(x=t, y=f, color=method),
             show.legend=F) +
  geom_line(data=rbind(results_sparse, results_dense, results_true),
            aes(x=t, y=f, color=method)) +
  scale_color_brewer(palette = "Dark2") + 
  scale_fill_brewer(palette="Dark2") +
  ylab('I(t)')
```

This shows that a step size of $\Delta t=1$ day results in a very inaccurate solution. In particular, because the derivative is increasing
over time, the forward Euler method (which, between solver time steps, assumes
that the  derivative remains constant at an old value) produces an approximate
solution that is too low. For $\Delta t=0.1$ day, the
approximate solution is more accurate, although still significantly understates the true solution.


## Effect of integrator step size on the SEIRD model
We now examine the impact of step size on the `SEIRD` model, where we set its parameter values to be representative of the Delta variant of SARS-CoV-2. First, we load the `SEIRD` model from comomodels and specify plausible parameter values and initial conditions.

```{r}
# instantiate model
model <- comomodels::SEIRD()

# delta parameters
params <- covid_transmission_parameters(variant="delta")
kappa <- params$kappa
mu <- params$mu
gamma <- params$gamma
R0_target <- params$R0
# set beta to hit R0 target
beta <- (mu + gamma) * R0_target

transmission_parameters(model) <- list(beta=beta, kappa=kappa, gamma=gamma, mu=mu)

# assume small proportion initially infected
initial_conditions(model) <- list(S0=0.999, E0=0, I0=0.001, R0=0)
```

Next, we solve the model using the forward Euler method. We investigate two different choices for the solver step size. The first is a daily time step,
which naively might be considered an acceptable default choice, given that we 
are interested in simulating daily cases and deaths. We also consider a much smaller time step of 0.001 days, for which we know that the solution is 
reasonably accurate (see the section "Adaptive step size methods" below for further details on obtaining numerical solutions with a desired degree of 
accuracy). We then plot the daily deaths and cases below.

```{r, fig.width=11, fig.height=5}
times_daily <- seq(0, 50, by=1)
solution_daily <- run(model, times_daily, solve_method="euler")
changes_daily <- solution_daily$changes

times_dense <- seq(0, 50, by=0.001)
solution_dense <- run(model, times_dense, solve_method="euler")
df <- solution_dense$changes
df <- as.data.frame(df)

# Sum the deaths and incidences from the dense output to obtain daily values
changes_dense <- data.frame(time=NULL, value=NULL, compartment=NULL)
for (t in times_daily) {
  cases <- subset(df, t-1<df[,1] & df[,1]<=t)
  incidence <- subset(cases, cases[,3]=="Incidence")
  death <- subset(cases, cases[,3]=="Deaths")
  changes_dense <- rbind(changes_dense, 
                         data.frame(time=t, 
                                    value=sum(incidence$value), 
                                    compartment="Incidence", 
                                    age_range="0-150"))
  changes_dense <- rbind(changes_dense, 
                         data.frame(time=t, 
                                    value=sum(death$value), 
                                    compartment="Deaths", 
                                    age_range="0-150"))
}

# Reorder the deaths and incidences
changes_dense <- changes_dense[order(changes_dense$compartment, changes_dense$time),]

# Save the labels and combine for plotting
changes_dense$method <- "dt=0.001 Day (accurate)"
changes_daily$method <- "dt=1 Day"
all_changes <- rbind(changes_dense, changes_daily)

ggplot(subset(all_changes, all_changes[,3]=="Deaths" | all_changes[,3]=="Incidence"), 
       aes(x = time, y = value, fill = method)) +
  geom_bar(stat="identity", position = position_dodge()) +
  facet_wrap(~ all_changes[,3], scales="free") +
  scale_color_brewer(palette = "Dark2") + 
  scale_fill_brewer(palette="Dark2")
```
These results indicate that for the `SEIRD` model with these parameters, a step size of one day results in substantial inaccuracy. This step size resulted in a higher peak case load which occurred later than the accurate solution. It also delayed the peak in deaths.

## Differential equations versus difference equations
The forward Euler recurrence relation defining the approximate solution to a
_differential_ equation,

$$f_{i+1} = f_i + \Delta t g(t_i, f_i),$$

is identical with a typical definition of a _difference_ equation of the sort used in multiple fields of computational biology, including epidemiology (see
[@izzo2009general]).

Thus, what we have considered as numerical error arising from the forward Euler method may alternatively be viewed as the discrepancy between an (accurately solved) differential equation model and a corresponding difference
equation model. The results above show that the difference in outputs from each approach can be nontrivial. While difference equations can be reasonable models for certain phenomena, our analysis shows that these models are not straightforward equivalents of differential equations. This is not surprising since differential equations suppose that individuals have a continuous overlap of generations, whereas difference equations suppose that these are non-overlapping [@murray1989mathematical]. As such, it is questionable whether difference equations should be used to model transmission dynamics of a disease such as COVID-19, where an infectious individual may "give birth" to a number of infected individuals who coexist with the infector.

## Adaptive step size methods

The chief advantage of numerical solvers such as forward Euler described above,
which take a fixed grid of time points on which to calculate the approximate
solution, is their ease of implementation. However, as seen above, these
methods suffer because:

1. To obtain an accurate solution, the solver step size must be set to some
small value, but it is often unknown how small this value must be.
2. When the solver step size is set to a small value, the method may be
impractically slow.

Both of these defects are addressed by adaptive step size methods. In these,
the user specifies not a step size but a _tolerance_—some relative or absolute
value which the local error in the approximate solution should not exceed. 
Then, the solver algorithm selects the appropriate step size in order for the
solution to meet this tolerance, using the theoretical error properties of the 
solver and various techniques for step size adaptation. Adaptive solvers are able to vary the step
size over the course of the solve, selecting very small values only in those
regions of time where this is necessary, such as where the solution is 
changing rapidly over time, and otherwise increasing the step size
to larger values. For this reason, they are much more efficient than fixed step
solvers [@gautschi2012numerical].

A wide variety of approaches to step size adaptation have been proposed. At 
each step, these methods typically use some estimator of the error
introduced at that time step, and then select roughly the largest possible step
such that the threshold imposed by the user-supplied tolerance is not exceeded.

Although adaptive step size solvers are challenging to implement, it is often
possible to rely on existing implementations. In R, a wide selection of 
adaptive step size solvers are provided by the deSolve package [@soetaert2010solving]. The comomodels
library uses deSolve to handle numerical solution of the systems of ODEs that 
arise in compartmental epidemiology models, giving users easy access to
efficient adaptive step size solvers. In comomodels, default values
of the local error tolerances are used, so these need not be specified
manually. The default solver used in comomodels is LSODA [@hindmarsh2005lsoda].
This method automatically switches between the Adams and backward
differentiation formula (BDF) solvers [@gautschi2012numerical] based on the
properties of the differential equation being solved.

In the following example, we compare the performance of the forward Euler 
solver with a fixed, small step size to that of the LSODA adaptive step size
solver.


```{r, fig.width=11, fig.height=5}
times_dense <- seq(0, 50, by=0.001)
print("Timing for Euler (fixed step):")
ptm <- proc.time()
solution_dense <- run(model, times_dense, solve_method='euler')
proc.time() - ptm

times_daily <- seq(0, 50, by=1)
print("Timing for LSODA (adaptive step):")
ptm <- proc.time()
solution_adapt <- run(model, times_daily, solve_method='lsoda')
proc.time() - ptm

df <- solution_dense$changes
df <- as.data.frame(df)

# Sum the deaths and incidences from the dense output to obtain daily values
changes_dense <- data.frame(time=NULL, value=NULL, compartment=NULL)
for (t in times_daily) {
  cases = subset(df, t-1<df[,1] & df[,1]<=t)
  incidence = subset(cases, cases[,3]=="Incidence")
  death = subset(cases, cases[,3]=="Deaths")
  changes_dense = rbind(changes_dense, 
                        data.frame(time=t,
                                   value=sum(incidence$value),
                                   compartment="Incidence",
                                   age_range="0-150"))
  changes_dense = rbind(changes_dense,
                        data.frame(time=t,
                                   value=sum(death$value),
                                   compartment="Deaths",
                                   age_range="0-150"))
}

# Save the labels and combine for plotting
changes_adapt <- solution_adapt$changes
changes_dense$method <- "Euler, dt=0.001 Day"
changes_adapt$method <- "LSODA (adaptive step)"
all_changes <- rbind(changes_dense, changes_adapt)

ggplot(subset(all_changes, all_changes[,3]=="Deaths" | all_changes[,3]=="Incidence"), 
       aes(x = time, y = value, fill = method)) +
  geom_bar(stat="identity", position = position_dodge()) +
  facet_wrap(~ all_changes[,3], scales="free") +
  scale_color_brewer(palette = "Dark2") + 
  scale_fill_brewer(palette="Dark2")

```
Both methods achieve the same accurate solution. However, for the LSODA solver,
it is not necessary to manually select a time step: we merely provide the time points at 
which we want to access the solution (daily), and the solver handles the
selection of the step sizes. We also note the significant speed advantage of
the LSODA solver, which may be more than 50 times faster (the actual runtimes
will be system dependent).

Adaptive solvers are not infallible. On highly challenging differential
equations, and when supplied with insufficient tolerances for the simulation or
inference task at hand, they may cause problems for simulation and inference. However,
for the simulation of compartmental epidemiological models such as those
covered by the comomodels package, adaptive solvers such as LSODA (the default choice of solver in
comomodels) generally perform better in terms of solution fidelity and speed.


## Discontinuities in the RHS (interventions)

Differential equations of the form

$$\frac{df}{dt} = g(t, f)$$

where $g$ is discontinuous present further challenges to both adaptive and fixed step size methods, whose error properties (and, for adaptive step size methods, the algorithms used to determine the time steps) typically make assumption such as differentiability and Lipschitz continuity of $g$.

In the epidemiological models of the sort described by comomodels, discontinuities of $g$ in time are possible, most notably in the SEIRDV model where step function interventions may be used to specify the rate of vaccination.

Some of the numerical methods that can be used to increase efficiency and decrease error in the case of discontinuous $g$ are:

1. Divide the time interval into sets where $g$ is continuous, and solve the ODE on each set separately.
2. Replace the discontinuous $g$ with a smooth approximation, and use a standard solver.

The second approach is adopted by the SEIRDV model, as a tanh approximation to the step function is used.

First, we study the effects of a discontinuity in the vaccination rate in the 
SEIRDV model without a tanh approximation. The vaccination rate is set to 0.15 from $t=4.75$ to $t=9.75$, and 0
elsewhere. The model is solved using an Euler solver with a large time step 
of 0.5 days, once without any consideration of the discontinuities (green curve) and once
with solver time steps inserted at the discontinuities (blue curve, method 1 above). These 
results are compared to an accurate (but potentially inefficient) solution
obtained using the LSODA solver. The value of the $I$ compartment is shown over time, and the time period of vaccination is shaded in gray.

```{r, fig.width=8, fig.height=5}
model <- SEIRDV()
# Select transmission parameters where the effect of the discontinuity can be 
# observed on a short time scale of 20 days
params <- list(beta=1, kappa=0.9, gamma=0.5, mu=0.1, nu=0.4, delta_V=0.15, delta_R=0.05, delta_VR = 0.02)
transmission_parameters(model) <- params
initial_conditions(model) <- list(S0=0.99, E0=0, I0=0.01, R0=0, V0=0, VR0=0)
intervention_parameters(model) <- list(starts=c(4.75),
                                       stops=c(9.75),
                                       coverages=c(1),
                                       tanh_slopes=1000)

times <- seq(0, 20, by = 0.5)
df1 <- run(model, times, solve_method="lsoda")$states
df1$method = "Accurate (LSODA)"

solver_times <- seq(0, 20, by=0.5)
df2 <- run(model, solver_times, solve_method="euler")$states
df2$method = "Euler, dt=0.5 day"


solver_times <- sort(c(solver_times, 4.75, 9.75))
df3 <- run(model, solver_times, solve_method="euler")$states
df3$method = "Euler, dt=0.5 day, start/stop at discon"

df <- rbind(df1, df3, df2)

ggplot(subset(df, df[,3]=="I"), 
       aes(x = time, y = value)) +
  geom_rect(aes(xmin=4.75,
                xmax=9.75,
                ymin = -Inf,
                ymax = Inf), fill = 'gray', alpha = 0.025) +
  geom_line(aes(color=method)) +
  ylab('I compartment')

```

Both Euler solvers experience large error due to the large step size. However,
additional error caused by the discontinuous $\delta_V$ is apparent, as, after the
vaccination period begins, the Euler method without consideration of the discontinuity
starts to deviate more drastically from the true solution.

An alternative approach is to smooth the discontinuity. In comomodels SEIRDV,
this can be performed by supplying an appropriate slope to the tanh function
that is used to smooth the $\delta_V$ profile:

```{r, fig.width=8, fig.height=5}
int_parms <- InterventionParameters(start=c(4.75),
                                    stop=c(9.75),
                                    coverage=c(1),
                                    tanh_slope=1000)

sim_parms <- SimulationParameters(start=0, stop=20, tstep = 0.1)
i1 <- intervention_protocol(int_parms, sim_parms)
i1$method <- "Original profile"

int_parms2 <- InterventionParameters(start=c(4.75),
                                    stop=c(9.75),
                                    coverage=c(1),
                                    tanh_slope=2)
sim_parms2 <- SimulationParameters(start=0, stop=20, tstep = 0.1)
i2 <- intervention_protocol(int_parms2, sim_parms2)
i2$method <- "tanh-smoothed profile"

ggplot(rbind(i1, i2), aes(x=time, y=coverage, color=method)) +
  geom_line() +
  scale_color_brewer(palette = "Dark2") +
  labs(x = "time (days)", y = "level of intervention")
```
Next, we compare the effects of the tanh smoothing on the simulated incidences.
The value of the $I$ compartment is shown over time, and the time period of vaccination is shaded in gray.

```{r, fig.width=8, fig.height=5}
model <- SEIRDV()
params <- list(beta=1, kappa=0.9, gamma=0.5, mu=0.1, nu=0.4, delta_V=0.15, delta_R=0.05, delta_VR=0.01)
transmission_parameters(model) <- params
initial_conditions(model) <- list(S0=0.99, E0=0, I0=0.01, R0=0, V0=0, VR0=0)
intervention_parameters(model) <- list(starts=c(4.75),
                                       stops=c(9.75),
                                       coverages=c(1),
                                       tanh_slopes=1000)

times <- seq(0, 20, by = 0.5)
df1 <- run(model, times, solve_method="lsoda")$states
df1$method = "Without tanh smoothing"


intervention_parameters(model) <- list(starts=c(4.75),
                                       stops=c(9.75),
                                       coverages=c(1),
                                       tanh_slopes=2)

times <- seq(0, 20, by = 1)
df2 <- run(model, times, solve_method="lsoda")$states
df2$method = "With tanh smoothing"

df <- rbind(df1, df2)

ggplot(subset(df, df[,3]=="I"), 
       aes(x = time, y = value)) +
  geom_rect(aes(xmin=4.75,
                xmax=9.75,
                ymin = -Inf,
                ymax = Inf), fill = 'gray', alpha = 0.025) +
  geom_line(aes(color=method)) +
  ylab('I compartment')
```
Some slight differences are observed as a result of the tanh smoothing. Although the
tanh smoothing does change the model behavior, in most applications of epidemiological
modelling the smoothed version of the model is not only more tractable for numerical
solvers but also more closely resembles realistic situations, where policies and
interventions that change the parameters of the model take some time to be implemented.

Further details and a general treatment may be found in [@stewart2011dynamics], Chapter 8.


## References
