---
title: "Interventions of age-structured SEIRD"
output: rmarkdown::github_document
bibliography: ../references.bib
vignette: >
  %\VignetteIndexEntry{Interventions of age-structured SEIRD}
  %\VignetteEncoding{UTF-8}
  %\VignetteEngine{knitr::rmarkdown}
editor_options:
  markdown:
    wrap: 72
---

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>"
)
```

```{r setup}
library(comomodels)
library(deSolve)
library(ggplot2)
library(tidyverse)
library(glue)
```

## Introduction

In order to minimise the effects of an epidemic on the population, different types
of measures can introduced to achieve this. These interventions can be either
non-pharmaceutical in nature (e.g.\ lockdowns), or can dependent on new discoveries
in the medical field such as the use of antiviral treatments or the implementation
of vaccination campaigns.

However, within a population not all individuals are equally vulnerable. Older age
groups have overall in the case of the Covid epidemic a higher mortality rate than
their younger counterparts. Since these interventions take some time from the moment
they are introduced to produce a visible effect (e.g. in the case of vaccination,
not all the population can be immunised in a very short time span), priority is
hence given to those in the more vulnerable categories (e.g. vaccinate older people first).

We can integrate these interventions in the basic age-structured SEIRD model
framework through small alterations to the system of ODEs. In the case of vaccinations
